---
title: "California Policing"
description: "Using SQL to compare policing statistics"
author: Fox Kloske
date: December 1, 2025
format: html
title-block-banner: "media/CA_highway.png"
title-block-banner-color: white

#Remove messages/warnings
code-fold: true
execute: 
  warning: false
  message: false
---

The Stanford Open Policing Project of nearly 100 million traffic stops across the United States. This page uses California data to think critically about how different areas police.

```{r}
con_traffic <- DBI::dbConnect(

  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```

A note: understanding policing goes far beyond statistics. For example, community trust and relationships are difficult to quantify. The following is not meant to be comprehensive, but descriptive. For this project, I'd like to think about how traffic stops escalate. First, for the cities with available data, how many traffic stops result in a search? 

```{sql, connection = con_traffic, output.var = "temp_output"}

SELECT
  'Oakland' AS PoliceDepartment,
  COUNT(*) AS numStops,
  SUM(search_conducted) AS numSearches,
  AVG(search_conducted) AS pctSearches,
  COUNT(*) / 440983 AS searchesByPop
FROM ca_oakland_2020_04_01

UNION

SELECT
  'San Diego' AS PoliceDepartment,
  COUNT(*) AS numStops,
  SUM(search_conducted) AS numSearches,
  AVG(search_conducted) AS pctSearches,
  COUNT(*) / 1386932 AS searchesByPop
FROM ca_san_diego_2020_04_01

UNION

SELECT
  'San Francisco' AS PoliceDepartment,
  COUNT(*) AS numStops,
  SUM(search_conducted) AS numSearches,
  AVG(search_conducted) AS pctSearches,
  COUNT(*) / 827526 AS searchesByPop
FROM ca_san_francisco_2020_04_01

UNION

SELECT
  'San Jose' AS PoliceDepartment,
  COUNT(*) AS numStops,
  SUM(search_conducted) AS numSearches,
  AVG(search_conducted) AS pctSearches,
  COUNT(*) / 997368 AS searchesByPop
FROM ca_san_jose_2020_04_01

UNION

SELECT
  'Stockton' AS PoliceDepartment,
  COUNT(*) AS numStops,
  SUM(search_conducted) AS numSearches,
  AVG(search_conducted) AS pctSearches,
  COUNT(*) / 324975 AS searchesByPop
FROM ca_stockton_2020_04_01;
```


```{r}
library(ggplot2)
 library(dplyr)

temp_output <- temp_output |>
  mutate(searchesByPop = searchesByPop*100) |>
  mutate(pctSearches = pctSearches*100)

names(temp_output) <- 
  c("Police_Department", "Number_of_Stops", "Number_of_Searches", "Percentage_of_Searches", "Stops_by_Population (%)")

temp_output
```

TKTK

```{r}

temp_output |> 
  ggplot(aes(x = Police_Department, y = Percentage_of_Searches, fill = Police_Department)) +
  geom_col() + 
  labs(
    title = "Percentage of Stops that Lead to a Search",
    x = "Police Department",
    y = "Search Rate (%)") + 
    theme(legend.position = "none")
```

TKTK

```{sql, connection = con_traffic, output.var = "temp_output2"}

#WHERE does not work inside SUM, instead this new syntax: CASE WHEN, THEN, ELSE, END
SELECT
  'Oakland' AS PoliceDepartment,
  SUM(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitation,
  SUM(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrest,
  AVG(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitationPercent,
  AVG(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrestPercent
FROM ca_oakland_2020_04_01

UNION

SELECT
  'San Diego' AS PoliceDepartment,
  SUM(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitation,
  SUM(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrest,
  AVG(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitationPercent,
  AVG(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrestPercent
FROM ca_san_diego_2020_04_01

UNION

SELECT
  'San Francisco' AS PoliceDepartment,
  SUM(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitation,
  SUM(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrest,
  AVG(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitationPercent,
  AVG(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrestPercent
FROM ca_san_francisco_2020_04_01

UNION

SELECT
  'San Jose' AS PoliceDepartment,
  SUM(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitation,
  SUM(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrest,
  AVG(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitationPercent,
  AVG(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrestPercent
FROM ca_san_jose_2020_04_01

UNION

SELECT
  'Stockton' AS PoliceDepartment,
  SUM(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitation,
  SUM(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrest,
  AVG(CASE WHEN search_conducted = 1 AND citation_issued = 1 THEN 1 ELSE 0 END) AS searchToCitationPercent,
  AVG(CASE WHEN search_conducted = 1 AND arrest_made = 1 THEN 1 ELSE 0 END) AS searchToArrestPercent
FROM ca_stockton_2020_04_01;
```

```{r}
library(tidyverse)

temp_output2 |> 
  mutate(searchToCitationPercent = searchToCitationPercent*100, searchToArrestPercent = searchToArrestPercent*100) |>
  #Wow, I new pivots would come in handy!
  pivot_longer(
    cols = c(searchToCitationPercent, searchToArrestPercent),
    names_to = "searchResult",
    values_to = "resultPercent") |>
  ggplot(aes(x = PoliceDepartment, y = resultPercent, fill = searchResult)) +
  #Dodge staggers the percents
  geom_col(position="dodge") + 
  labs(
    title = "Percentage of Stops that Lead to a Search",
    x = "Police Department",
    y = "Search Rate (%)") + 
    theme(legend.position = "none")
```

TKTK

```{r}
temp_output2

DBI::dbDisconnect(con_traffic)
```
